---
title: "BMEG 310 Final Project"
author: "Justin Yoon, Zeena ..., Addin ..."
date: "2024-11-28"
output: html_document
---

https://github.com/jyoon1325/BMEG310FinalProject.git

Import libraries:
```{r}
#all libraries from tutorials:

#Introduction to R Basics
library(graphics)
#install.packages("ggplot2")
library(ggplot2)
#install.packages("RColorBrewer")
library(RColorBrewer)
#install.packages("rmarkdown")
library(rmarkdown)

#Introduction to Machine Learning
#install.packages("devtools")
library(devtools)
#install.packages("remotes")
#remotes::install_github("vqv/ggbiplot")
library(ggbiplot)
suppressPackageStartupMessages(library(dplyr))
#install.packages("readxl")
library(readxl)#read_excel
#install.packages("ISLR")
require(ISLR)
#install.packages("corrplot")
library(corrplot)
#install.packages("caret")
library(caret)
install.packages("randomForest")
library(randomForest)

#Introduction to Survival Analysis
#BiocManager::install("TCGAbiolinks")
library("TCGAbiolinks")
#BiocManager::install("survival")
library("survival")
#BiocManager::install("survminer")
library("survminer")
library("SummarizedExperiment")

#Introduction to Mutation Analysis
# install.packages("pheatmap")
library(pheatmap)

#Introduction to Differential Analysis
# BiocManager::install("AnnotationDbi")
# BiocManager::install("org.Hs.eg.db")
# BiocManager::install("pathview")
# BiocManager::install("gage")
# BiocManager::install("gageData")
library("gridExtra")
library(ggrepel)
library("AnnotationDbi")
library("org.Hs.eg.db")
library(pathview)
library(gage)
library(gageData)


#Additional useful libraries:
library(tidyr)
library(readr)
library(data.table)
library(stats)
library(cluster)
library(factoextra)
library(org.Hs.eg.db)
library(enrichplot)
```



Import data:
```{r}
RNAseq_LIHC <- read.csv("RNAseq_LIHC.csv", header = TRUE, sep = "\t")
data_clinical_patient <- read.delim("data_clinical_patient.txt", header = TRUE, sep = "\t")
data_mutations <- read.delim("data_mutations.txt", header = TRUE, sep = "\t")
```

summaries of data
```{r}
#summary(RNAseq_LIHC)
#summary(data_clinical_patient)
#summary(data_mutations)
```


Analysis:
1. Clinical Data
2. Mutation Data
3. Expression Data


Clinical Data:
1. Explore clinical data wrt the distribution based on various variable/attributes such as age, stage, survival, etc 

Age Distribution
```{r}
#convert Diagnosis Age to numeric
data_clinical_patient$Diagnosis.Age <- as.numeric(data_clinical_patient$Diagnosis.Age)

#plot in a bar graph
ggplot(data_clinical_patient, aes(x = Diagnosis.Age)) +
  geom_histogram(fill = "blue", color = "black") +
  labs(title = "Distribution of Diagnosis Age", x = "Age", y = "Frequency") +
  theme_minimal()
```
cancer stage Distribution
```{r}
ggplot(data_clinical_patient, aes(x = Neoplasm.Disease.Stage.American.Joint.Committee.on.Cancer.Code)) +
  geom_bar(fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Cancer Stages", x = "Cancer Stage", y = "Frequency") +
  theme_minimal()
```
Survival Status Distribution
```{r}
table(data_clinical_patient$Overall.Survival.Status)

#count survival status
survival_status <- as.data.frame(table(data_clinical_patient$Overall.Survival.Status))
colnames(survival_status) <- c("Overall.Survival.Status", "Count")

#plot survival status distribution
ggplot(survival_status, aes(x = "", y = Count, fill = Overall.Survival.Status)) +
  geom_bar(stat = "identity") +
  coord_polar("y") +
  labs(title = "Overall Survival Status Distribution")
```
Diagnosis Age by Survival Status
```{r}
ggplot(data_clinical_patient, aes(x = as.numeric(Diagnosis.Age), fill = Overall.Survival.Status)) +
  geom_histogram() +
  labs(title = "Age Distribution by Survival Status", x = "Age", y = "Frequency") +
  theme_minimal()
```
Diagnosis Age by Cancer Stage
```{r}
ggplot(data_clinical_patient, aes(x = as.numeric(Diagnosis.Age), fill = Neoplasm.Disease.Stage.American.Joint.Committee.on.Cancer.Code)) +
  geom_histogram() +
  labs(title = "Age Distribution by Cancer Stage", x = "Age", y = "Frequency") +
  theme_minimal()
```
Survival Analysis
```{r}
# Check the structure and unique values in the survival time column
str(data_clinical_patient$Overall.Survival..Months.)
unique(data_clinical_patient$Overall.Survival..Months.)

# Convert 'Overall.Survival..Months.' to numeric, handling non-numeric values
data_clinical_patient$Overall.Survival..Months. <- as.numeric(as.character(data_clinical_patient$Overall.Survival..Months.))

# Check for and handle NAs (w/o it, problems occur later on)
if (sum(is.na(data_clinical_patient$Overall.Survival..Months.)) > 0) {
  data_clinical_patient$Overall.Survival..Months.[is.na(data_clinical_patient$Overall.Survival..Months.)] <- 
    median(data_clinical_patient$Overall.Survival..Months., na.rm = TRUE)
}

# summary(data_clinical_patient$Overall.Survival..Months.)

#Create Survival Object
surv_obj <- Surv(
  time = data_clinical_patient$Overall.Survival..Months.,
  event = data_clinical_patient$Overall.Survival.Status == "Deceased"
)

#inspect
#str(surv_obj)

#Kaplan-Meier Survival Curve
# Fit Kaplan-Meier survival curve
fit <- survfit(surv_obj ~ 1, data = data_clinical_patient)

#Plot Kaplan-Meier survival curve
ggsurvplot(
  fit, data = data_clinical_patient,
  title = "Kaplan-Meier Survival Curve",
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  conf.int = TRUE,   # Show confidence intervals
  risk.table = TRUE, # Show the risk table
  ggtheme = theme_minimal()
)

#Stratify by Cancer Stage
# Fit survival curve by cancer stage
fit_stage <- survfit(surv_obj ~ Neoplasm.Disease.Stage.American.Joint.Committee.on.Cancer.Code, data = data_clinical_patient)

# Plot stratified survival curves
ggsurvplot(
  fit_stage, data = data_clinical_patient,
  title = "Kaplan-Meier Survival Curve by Cancer Stage",
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  conf.int = TRUE,
  risk.table = TRUE,
  legend.title = "Cancer Stage",
  ggtheme = theme_minimal()
)

#Statistical Test - Log-Rank Test
#log_rank_test <- survdiff(surv_obj ~ Neoplasm.Disease.Stage.American.Joint.Committee.on.Cancer.Code, data = data_clinical_patient)
#print(log_rank_test)

#Data Quality Histogram of Survival Times
hist(
  data_clinical_patient$Overall.Survival..Months.,
  main = "Survival Time Distribution",
  xlab = "Time (Months)",
  breaks = 30,
  col = "skyblue"
)
```

Mutation data:
1. Build a gene-patient matrix that contains mutation data. Within this matrix, "1" represents whether there is a mutation in a given gene and given patient and "0" represents no mutation. Note, when building this matrix, you can take into account the type of mutations (e.g., synonymous, non-synonymous, etc) 
2. Pick the top 20 or most frequently mutated genes from this matrix.  
3. Perform clustering on the resulting mutation matrix Perhaps try ward and ward.D2 linkage methods if you see narrow clusters.
4. Take the clusters of patients and do (a) DE analysis on all genes (not just top 500), (b) do pathway analysis on the DE genes that pass the adjusted p-value threshold (not just the top 20 genes) to see if specific pathways are hit, (c) see if outcome data between patient clusters is different, (d) see if any other clinical data is different when you compare patient clusters, (e) explore the correlation between the patient clusters that you have got from mutation and expression data - how much overlap do you see between the patient clusters?
```{r}

```

Expression data:
1. Normalize expression data using DEseq2. This was explained and included in the lab document. 
2. Pick the top n (e.g., 500, 1000, ...) most variable genes. This could be done before step 1 or after step 1. 
3. Perform clustering on the resulting expression matrix. Perhaps try ward and ward.D2 linkage methods if you see narrow clusters.
4. Take the clusters of patients and do (a) DE analysis on all genes (not just top 500), (b) do pathway analysis on the DE genes that pass the adjusted p-value threshold (not just the top 20 genes) to see if specific pathways are hit, (c) see if outcome data between the patient clusters is different, (d) see if any other clinical data is different when you compare patient clusters, (e) see if patient clusters are enriched with any type of mutation data.   
```{r}

```


