---
title: "BMEG 310 Final Project"
author: "Justin Yoon, Zeena ..., Addin ..."
date: "2024-11-28"
output: html_document
---

https://github.com/jyoon1325/BMEG310FinalProject.git

#Import libraries:
```{r}
#all libraries from tutorials:

#Introduction to R Basics
library(graphics)
#install.packages("ggplot2")
library(ggplot2)
#install.packages("RColorBrewer")
library(RColorBrewer)
#install.packages("rmarkdown")
library(rmarkdown)

#Introduction to Machine Learning
#install.packages("devtools")
library(devtools)
#install.packages("remotes")
#remotes::install_github("vqv/ggbiplot")
library(ggbiplot)
library(dplyr)
#install.packages("readxl")
library(readxl)#read_excel
#install.packages("ISLR")
require(ISLR)
#install.packages("corrplot")
library(corrplot)
#install.packages("caret")
library(caret)
#install.packages("randomForest")
library(randomForest)

#Introduction to Survival Analysis
#BiocManager::install("TCGAbiolinks")
library(TCGAbiolinks)
#BiocManager::install("survival")
library(survival)
#BiocManager::install("survminer")
library(survminer)
library(SummarizedExperiment)

#Introduction to Mutation Analysis
# install.packages("pheatmap")
library(pheatmap)

#Introduction to Differential Analysis
# BiocManager::install("AnnotationDbi")
# BiocManager::install("org.Hs.eg.db")
# BiocManager::install("pathview")
# BiocManager::install("gage")
# BiocManager::install("gageData")
library(gridExtra)
library(ggrepel)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(pathview)
#install.packages("gage")
library(gage)
#install.packages("gageData")
library(gageData)


#Additional useful libraries:
library(tidyr)
library(readr)
library(data.table)
library(stats)
library(cluster)
library(factoextra)
library(org.Hs.eg.db)
library(enrichplot)
library(DESeq2)  
#install.packages(magrittr)
library (dplyr)
library(TCGAbiolinks)
#BiocManager::install("edgeR")
library(edgeR)
#BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)
library(circlize)

```


#Import data:
```{r}
RNA.data <- read.csv("RNAseq_LIHC.csv")
patient.data <- read.table("data_clinical_patient.txt", header = TRUE, sep = "\t")
mutations.data <- read.table("data_mutations.txt", header = TRUE, sep = "\t")
```

##summaries of data
```{r}
#summary(RNA.data)
#summary(patient.data)
#summary(mutations.data)
#colnames(RNA.data)
#colnames(patient.data)
#colnames(mutations.data)
```


#filter data
```{r}
patient_df<-patient.data
patient_df$PATIENT_ID<-substr(patient_df$PATIENT_ID,9,12)

mutations_df<-mutations.data
mutations_df$Tumor_Sample_Barcode<-substr(mutations_df$Tumor_Sample_Barcode,9,12)

#make copy of RNA data frame without the first column
RNA_df <- RNA.data[,-1]

#rename columns to match patient id naming convention of other data sets
colnames(RNA_df)<-substr(colnames(RNA_df),9,12)

#find unique patients
unique_patients<-unique(patient_df$PATIENT_ID)
unique_mutations<-unique(mutations_df$Tumor_Sample_Barcode)
unique_RNA<-unique(colnames(RNA_df))
#find common patients
common_patients<-Reduce(intersect, list(unique_patients,unique_mutations,unique_RNA))

#patient data with patients in all three data sets
unq_patient_df<-patient_df[patient_df$PATIENT_ID %in% common_patients,]

#mutation data with patients in all three data sets
unq_mutations_df<-mutations_df[mutations_df$Tumor_Sample_Barcode %in% common_patients,]

#RNA data with patients in all three data sets
unq_RNA_ID<-RNA_df[, colnames(RNA_df)%in% common_patients]

#add back the first column (X) from the RNA data frame
RNA_df$X<-RNA.data$X[match(rownames(RNA_df), rownames(RNA.data))] 
 

#Take out Low impact mutations, intron and synonymous variants
filtered_mutations <- subset(unq_mutations_df,
                             ((unq_mutations_df$IMPACT != "LOW") & (unq_mutations_df$Consequence != "intron_variant") 
                              & (unq_mutations_df$Consequence != "synonymous_variant")))

```
##Summaries of filtered data
```{r}
#summary(unq_patient_df)
#summary(filtered_mutations)
#summary(RNA_df)

```


_______________________________________________________________________________________________________________________________________

Analysis:
1. Clinical Data
2. Mutation Data
3. Expression Data

_______________________________________________________________________________________________________________________________________




#Clinical Data Analysis:
_______________________________________________________________________________________________________________________________________
Clinical Data:
1. Explore clinical data wrt the distribution based on various variable/attributes such as age, stage, survival, etc 
_______________________________________________________________________________________________________________________________________

###Age Distribution
```{r}
ggplot(unq_patient_df, aes(x = AGE)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Age Distribution", x = "Age", y = "Count") +
  theme_minimal()
```


###Age vs Overall Survival
```{r}
ggplot(unq_patient_df, aes(x = AGE, fill = OS_STATUS)) +
  geom_histogram(binwidth = 5, position = "dodge", alpha = 0.7) +
  labs(title = "Age Distribution by Overall Survival", x = "Age", y = "Count", fill = "Overall Survival") +
  theme_minimal()
```
###Age vs Progression-Free Survival (PFS) Status
```{r}
ggplot(unq_patient_df, aes(x = AGE, fill = PFS_STATUS)) +
  geom_histogram(binwidth = 5, position = "dodge", alpha = 0.7) +
  labs(title = "Age Distribution by Progression-Free Survival Status", x = "Age", y = "Count", fill = "Progression-Free Survival Status") +
  theme_minimal()
```
###Age Distribution with Boxplot Grouped by Tumor Stage
```{r}
ggplot(unq_patient_df, aes(x = AJCC_PATHOLOGIC_TUMOR_STAGE, y = AGE, fill = AJCC_PATHOLOGIC_TUMOR_STAGE)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Age Distribution by Tumor Stage", x = "Tumor Stage", y = "Age") +
  theme_minimal() +
  theme(legend.position = "none")
```
###Age Distribution by Disease-Specific Survival Status
```{r}
ggplot(unq_patient_df, aes(x = AGE, fill = DSS_STATUS)) +
  geom_histogram(binwidth = 5, position = "stack", alpha = 0.7) +
  labs(title = "Age Distribution by Disease-Specific Survival Status", x = "Age", y = "Count", fill = "Disease-Specific Survival Status") +
  theme_minimal()
```

#Survival analysis

##Prepare Data
```{r}
#survival data
unq_patient_df <- unq_patient_df %>%
  mutate(
    #convert OS_STATUS to a logical deceased indicator
    deceased = ifelse(OS_STATUS == "1:DECEASED", TRUE, FALSE),
    #convert OS_MONTHS to numeric
    overall_survival = as.numeric(OS_MONTHS)
  )
```

##Kaplan-Meier Analysis by Gender
```{r}
#fit survival model for gender
fit_gender <- survfit(Surv(overall_survival, deceased) ~ SEX, data = unq_patient_df)

#Kaplan-Meier plot for gender
ggsurvplot(fit_gender, data = unq_patient_df, pval = TRUE, risk.table = TRUE, risk.table.col = "strata", risk.table.height = 0.3)
```

##Kaplan-Meier Analysis by Weight
```{r}
# categorize patients by weight
unq_patient_df <- unq_patient_df %>%
  mutate(weight_group = case_when(
    WEIGHT < 50 ~ "Underweight",
    WEIGHT >= 50 & WEIGHT < 70 ~ "Normal weight",
    WEIGHT >= 70 & WEIGHT < 90 ~ "Overweight",
    WEIGHT >= 90 ~ "Obese",
    is.na(WEIGHT) ~ "Unknown"
  ))

#exclude patients with unknown weight
unq_patient_df_filtered <- unq_patient_df %>%
  filter(weight_group != "Unknown")

#fit survival model for weight groups
fit_weight <- survfit(Surv(overall_survival, deceased) ~ weight_group, data = unq_patient_df_filtered)

#Kaplan-Meier for weight groups
ggsurvplot(fit_weight, 
           data = unq_patient_df_filtered, 
           pval = TRUE,                        # Display p-value for log-rank test
           risk.table = TRUE,                  # Add risk table below the plot
           risk.table.col = "strata",          # Color risk table by strata
           risk.table.height = 0.3,            # Adjust risk table height
           title = "Kaplan-Meier Survival Analysis by Weight Group",
           xlab = "Time (Months)", 
           ylab = "Survival Probability")

```


##Kaplan-Meier Analysis by Tumor Stage
```{r}
# remove sub-stages of AJCC_PATHOLOGIC_TUMOR_STAGE
unq_patient_df <- unq_patient_df %>%
  mutate(tumor_stage = gsub("[abc]$", "", AJCC_PATHOLOGIC_TUMOR_STAGE))

#fit survival model by tumor stage
fit_stage <- survfit(Surv(overall_survival, deceased) ~ tumor_stage, data = unq_patient_df)

#Kaplan-Meier for tumor stage
ggsurvplot(fit_stage, 
           data = unq_patient_df, 
           pval = TRUE, 
           risk.table = TRUE, 
           risk.table.col = "strata", 
           risk.table.height = 0.3,
           title = "Survival by Tumor Stage")

```
##Kaplan-Meier Analysis by race
```{r}
#Group races 
  #"white", "asian", "black", and "others"...?
unq_patient_df <- unq_patient_df %>%
  mutate(race_group = ifelse(RACE %in% c("white", "asian", "black or african american"), RACE, "others"))

#fit survival model by race group
fit_race <- survfit(Surv(overall_survival, deceased) ~ race_group, data = unq_patient_df)

#Kaplan-Meier plot for race
ggsurvplot(fit_race, 
           data = unq_patient_df, 
           pval = TRUE, 
           risk.table = TRUE, 
           risk.table.col = "strata", 
           risk.table.height = 0.3,
           title = "Survival by Race Group")

```
##Kaplan-Meier Analysis by Age Group
```{r}
# age groups
unq_patient_df <- unq_patient_df %>%
  mutate(age_group = case_when(
    AGE < 50 ~ "<50",
    AGE >= 50 & AGE < 65 ~ "50-65",
    AGE >= 65 ~ "65+"
  ))

#fit survival model by age group
fit_age <- survfit(Surv(overall_survival, deceased) ~ age_group, data = unq_patient_df)

#Kaplan-Meier
ggsurvplot(fit_age, 
           data = unq_patient_df, 
           pval = TRUE, 
           risk.table = TRUE, 
           risk.table.col = "strata", 
           risk.table.height = 0.3,
           title = "Survival by Age Group")

```









#Create Clusters Using Expression Data

### Pre-Processing 

```{r}
#Move Column "X" to the first position

RNA_df <- RNA_df[, c("X", setdiff(names(RNA_df), "X"))]

#remove column "X", as it is not a sample
preprocessed_RNA = RNA_df[,-1]

#convert to matrix
preprocessed_RNA = as.matrix(preprocessed_RNA) 

gene_names <- RNA_df$X  # Extract gene names
rownames(preprocessed_RNA) <- gene_names  # Set as row names

#filter data that has 0 or 1 read count across all samples

preprocessed_RNA <- preprocessed_RNA[rowSums(preprocessed_RNA)>1,]
```

```{r}
# CPM normalize data
cpm_RNA <- cpm(preprocessed_RNA)

#log normalize the counts
log_RNA <- log2(cpm_RNA + 1)

```

```{r}
#Find the 50 most variably expressed genes

#Find Variance for each gene
expression_variance <- apply(log_RNA, 1, var)

#Sort based on variance
sorted_expression <- sort(expression_variance, decreasing = TRUE)

#Get top 50 genes
top_genes <- names(sorted_expression)[1:50]

```

```{r}
#subset the expression data to only include the top 50 genes
top_RNA <- log_RNA[top_genes, ]

#Visualize using a heatmap
pheatmap(top_RNA, scale = "row", clustering_distance_rows = "euclidean", 
         clustering_distance_cols = "euclidean", main = "Top 50 Variable Genes")
```

### Clustering 

```{r}
# Create a distance matrix
# Transpose data to cluster
dist_matrix <- dist(t(top_RNA), method = "euclidean")

#Hierarchical clustering
hc <- hclust(dist_matrix, method = "ward.D2") #ward's method 

#plot
plot(hc, main = "Hierarchical Clustering of Samples", xlab = "", sub = "")
```
```{r}
#Cut the Dendrogram into Clusters
k <- 2
clusters <- as.data.frame(cutree(hc, k=k))
colnames(clusters) <- "Cluster"

table(clusters)
```

```{r}
#PCA 
transposed_RNA <- t(top_RNA)  #transpose
pca <- prcomp(transposed_RNA, scale = TRUE)

pca_df <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], Cluster = factor(clusters$Cluster), PATIENT_ID = rownames(clusters))

#put in labels
pca_df$PATIENT_ID <- colnames(top_RNA)

#visualize with ggplot
ggplot(pca_df, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size =4) +
  labs(title = paste("PCA Plot with", k, "Clusters"), x = "PC1", y = "PC2") +
  theme_minimal()

```




#### Frequency of Gene Muation Comparison Between Clusters

```{r}
#split clustered patients into lists
cluster_pat <- split(rownames(clusters), clusters$Cluster)

clust1_pat <- cluster_pat[[1]] #cluster 1 patients
clust2_pat <- cluster_pat[[2]] #cluster 2 patients
```

```{r}
#split mutation data into sets based on patients in each cluster
clust1_mut <- filtered_mutations[filtered_mutations$Tumor_Sample_Barcode %in% clust1_pat,]
clust2_mut <- filtered_mutations[filtered_mutations$Tumor_Sample_Barcode %in% clust2_pat,]
```

```{r}

#Finding the top 20 most frequently mutated genes in Cluster 1

clust1_genes_freq <- as.data.frame(table(clust1_mut$Hugo_Symbol))

clust1_genes_ordered <- clust1_genes_freq[order(-clust1_genes_freq$Freq),]

ggplot(data = clust1_genes_ordered[1:20,], aes(x = Var1, y = Freq)) +
  geom_col()+ #bar chart 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+ #tilt text to make legible
  scale_x_discrete( limits = clust1_genes_ordered[1:20,]$Var1) #to retain new order
```

```{r}

#Finding the top 20 most frequently mutated genes in Cluster 2

clust2_genes_freq <- as.data.frame(table(clust2_mut$Hugo_Symbol))

clust2_genes_ordered <- clust2_genes_freq[order(-clust2_genes_freq$Freq),]

ggplot(data = clust2_genes_ordered[1:20,], aes(x = Var1, y = Freq)) +
  geom_col()+ #bar chart 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+ #tilt text to make legible
  scale_x_discrete( limits = clust2_genes_ordered[1:20,]$Var1) #to retain new order
```
Find whether there is a meaningful difference in genes mutated between the clusters. 

```{r}
# Normalize Mutation Frequencies (to account for difference in cluster sizes)

# Normalize mutation frequencies for cluster 1
clust1_genes_ordered$Normalized_Freq <- 
  clust1_genes_ordered$Freq / 266

# Normalize mutation frequencies for cluster 2
clust2_genes_ordered$Normalized_Freq <- 
  clust2_genes_ordered$Freq / 105

```

```{r}

#combine top genes from each cluster for comparison

# Add a cluster identifier to each data frame
clust1_genes_ordered$Cluster <- "Cluster 1"
clust2_genes_ordered$Cluster <- "Cluster 2"

# Combine the top 15 genes from both clusters
top_genes_combined <- rbind(clust1_genes_ordered[1:15,], clust2_genes_ordered[1:15,])

#plot
ggplot(top_genes_combined, aes(x = Var1, y = Normalized_Freq, fill = Cluster)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Normalized Mutation Frequencies by Cluster", 
       x = "Gene", y = "Normalized Frequency")

#we are interested in the genes with a significant mutational frequency difference between the two clusters
```
#### Repeat the Process for Chromosome

```{r}
# Frequency of mutations by chromosome in Cluster 1
clust1_chrom_freq <- as.data.frame(table(clust1_mut$Chromosome))

# Order chromosomes by frequency
clust1_chrom_ordered <- clust1_chrom_freq[order(-clust1_chrom_freq$Freq),]

# Plot the top 20 mutated chromosomes in Cluster 1
ggplot(data = clust1_chrom_ordered[1:20,], aes(x = Var1, y = Freq)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(limits = clust1_chrom_ordered[1:20,]$Var1) +
  labs(title = "Top 20 Mutated Chromosomes in Cluster 1", x = "Chromosome", y = "Frequency")
```
```{r}
# Frequency of mutations by chromosome in Cluster 2
clust2_chrom_freq <- as.data.frame(table(clust2_mut$Chromosome))

# Order chromosomes by frequency
clust2_chrom_ordered <- clust2_chrom_freq[order(-clust2_chrom_freq$Freq),]

# Plot the top 20 mutated chromosomes in Cluster 2
ggplot(data = clust2_chrom_ordered[1:20,], aes(x = Var1, y = Freq)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(limits = clust2_chrom_ordered[1:20,]$Var1) +
  labs(title = "Top 20 Mutated Chromosomes in Cluster 2", x = "Chromosome", y = "Frequency")

```

```{r}
# Normalize mutation frequencies for Cluster 1
clust1_chrom_ordered$Normalized_Freq <- clust1_chrom_ordered$Freq / 266

# Normalize mutation frequencies for Cluster 2
clust2_chrom_ordered$Normalized_Freq <- clust2_chrom_ordered$Freq / 105

# Add a cluster identifier to each dataframe
clust1_chrom_ordered$Cluster <- "Cluster 1"
clust2_chrom_ordered$Cluster <- "Cluster 2"

# Combine the top 15 chromosomes from both clusters
top_chrom_combined <- rbind(clust1_chrom_ordered[1:15,], clust2_chrom_ordered[1:15,])

# Plot the normalized mutation frequencies by cluster
ggplot(top_chrom_combined, aes(x = Var1, y = Normalized_Freq, fill = Cluster)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Normalized Mutation Frequencies by Chromosome", 
       x = "Chromosome", y = "Normalized Frequency")

```

#### Repeat the Process for Consequence 
(to compare the resultant types of variants)

```{r}
# Frequency of mutation consequences in Cluster 1
clust1_conseq_freq <- as.data.frame(table(clust1_mut$Consequence))

# Order consequences by frequency
clust1_conseq_ordered <- clust1_conseq_freq[order(-clust1_conseq_freq$Freq),]

# Plot the top 20 most frequent consequences in Cluster 1
ggplot(data = clust1_conseq_ordered[1:10,], aes(x = Var1, y = Freq)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(limits = clust1_conseq_ordered[1:10,]$Var1) +
  labs(title = "Top 20 Mutation Consequences in Cluster 1", x = "Consequence", y = "Frequency")

```

```{r}
# Frequency of mutation consequences in Cluster 2
clust2_conseq_freq <- as.data.frame(table(clust2_mut$Consequence))

# Order consequences by frequency
clust2_conseq_ordered <- clust2_conseq_freq[order(-clust2_conseq_freq$Freq),]

# Plot the top 20 most frequent consequences in Cluster 2
ggplot(data = clust2_conseq_ordered[1:20,], aes(x = Var1, y = Freq)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(limits = clust2_conseq_ordered[1:20,]$Var1) +
  labs(title = "Top 20 Mutation Consequences in Cluster 2", x = "Consequence", y = "Frequency")

```

```{r}
# Normalize mutation frequencies for Cluster 1
clust1_conseq_ordered$Normalized_Freq <- clust1_conseq_ordered$Freq / 266

# Normalize mutation frequencies for Cluster 2
clust2_conseq_ordered$Normalized_Freq <- clust2_conseq_ordered$Freq / 105

# Add a cluster identifier to each dataframe
clust1_conseq_ordered$Cluster <- "Cluster 1"
clust2_conseq_ordered$Cluster <- "Cluster 2"

# Combine the top 15 consequences from both clusters
top_conseq_combined <- rbind(clust1_conseq_ordered[1:10,], clust2_conseq_ordered[1:10,])

# Plot the normalized mutation frequencies by cluster
ggplot(top_conseq_combined, aes(x = Var1, y = Normalized_Freq, fill = Cluster)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Normalized Mutation Consequences by Cluster", 
       x = "Consequence", y = "Normalized Frequency")

```
Observation: Cluster one had more missense variants. Cluster 1 had more downstream gene variants, while Cluster 2 had more upstream gene variants.










```{r}
# Extract List of patients in each cluster

clust1_IDs <- names(clusters[clusters == 1])
clust2_IDs <- names(clusters[clusters == 2])

# Save patient names to lists
list_of_patients <- list(
  Cluster1 = clust1_IDs,
  Cluster2 = clust2_IDs
)
```



#Survival Analysis on Clinical Data to compare Patients in the Clusters
####Kaplan-Meier Analysis of Clusters by Survival
```{r}
#add cluster information to the clinical data
unq_patient_df <- unq_patient_df %>%
  mutate(Cluster = clusters$Cluster[match(PATIENT_ID, rownames(clusters))])

#missing clusters?
unq_patient_df <- unq_patient_df %>%
  filter(!is.na(Cluster))

#survival model based on clusters
fit_clusters <- survfit(Surv(overall_survival, deceased) ~ Cluster, data = unq_patient_df)

#Kaplan-Meier Analysis of Clusters by Survival
ggsurvplot(
  fit_clusters, 
  data = unq_patient_df, 
  pval = TRUE,                        
  risk.table = TRUE,                  
  risk.table.col = "strata",          
  risk.table.height = 0.3,            
  title = "Kaplan-Meier Analysis of Clusters by Survival",
  xlab = "Time", 
  ylab = "Survival Probability",
)
```

####Kaplan-Meier Analysis of Clusters by Gender
```{r}
#survival model by gender within clusters
fit_gender_cluster <- survfit(Surv(overall_survival, deceased) ~ SEX + Cluster, data = unq_patient_df)

#Kaplan-Meier plot
ggsurvplot(
  fit_gender_cluster, 
  data = unq_patient_df, 
  pval = TRUE, 
  risk.table = TRUE, 
  risk.table.col = "strata", 
  risk.table.height = 0.3,
  facet.by = "Cluster",               #separate plots for each cluster (hehe looks pretty)
  title = "Kaplan-Meier Analysis of Clusters by Gender",
  xlab = "Time", 
  ylab = "Survival Probability"
)
```

####Kaplan-Meier Analysis of Clusters by Tumor stage
```{r}
#Simplify tumor stage for analysis
unq_patient_df <- unq_patient_df %>%
  mutate(tumor_stage = gsub("[abc]$", "", AJCC_PATHOLOGIC_TUMOR_STAGE))

#survival model by tumor stage within clusters
fit_tumor_stage_cluster <- survfit(Surv(overall_survival, deceased) ~ tumor_stage + Cluster, data = unq_patient_df)

#kaplan-Meier plot
ggsurvplot(
  fit_tumor_stage_cluster, 
  data = unq_patient_df, 
  pval = TRUE, 
  risk.table = TRUE, 
  risk.table.col = "strata", 
  risk.table.height = 0.3,
  facet.by = "Cluster",
  title = "Kaplan-Meier Analysis of Clusters by Tumor stage",
  xlab = "Time", 
  ylab = "Survival Probability"
)

```
####Kaplan-Meier Analysis of Clusters by Race
```{r}
#Simplify Group races
unq_patient_df <- unq_patient_df %>%
  mutate(race_group = ifelse(RACE %in% c("white", "asian", "black or african american"), RACE, "others"))

#survival model by race within clusters
fit_race_cluster <- survfit(Surv(overall_survival, deceased) ~ race_group + Cluster, data = unq_patient_df)

#Kaplan-Meier plot
ggsurvplot(
  fit_race_cluster, 
  data = unq_patient_df, 
  pval = TRUE, 
  risk.table = TRUE, 
  risk.table.col = "strata", 
  risk.table.height = 0.3,
  facet.by = "Cluster",
  title = "Kaplan-Meier Analysis of Clusters by Race",
  xlab = "Time", 
  ylab = "Survival Probability"
)

```

####Kaplan-Meier Analysis of Clusters by Age
```{r}
#categorize age groups
unq_patient_df <- unq_patient_df %>%
  mutate(age_group = case_when(
    AGE < 50 ~ "<50",
    AGE >= 50 & AGE < 65 ~ "50-65",
    AGE >= 65 ~ "65+"
  ))

#survival model by age group within clusters
fit_age_group_cluster <- survfit(Surv(overall_survival, deceased) ~ age_group + Cluster, data = unq_patient_df)

#Kaplan-Meier plot
ggsurvplot(
  fit_age_group_cluster, 
  data = unq_patient_df, 
  pval = TRUE, 
  risk.table = TRUE, 
  risk.table.col = "strata", 
  risk.table.height = 0.3,
  facet.by = "Cluster",
  title = "Kaplan-Meier Analysis of Clusters by Age",
  xlab = "Time", 
  ylab = "Survival Probability"
)

```


####Kaplan-Meier Analysis of Clusters by sex
```{r}
#model by SEX within clusters
fit_sex_cluster <- survfit(Surv(overall_survival, deceased) ~ SEX + Cluster, data = unq_patient_df)

# Kaplan-Meier plot
ggsurvplot(
  fit_sex_cluster, 
  data = unq_patient_df, 
  pval = TRUE, 
  risk.table = TRUE, 
  risk.table.col = "strata", 
  risk.table.height = 0.3,
  facet.by = "Cluster",
  title = "Kaplan-Meier Analysis of Clusters by sex",
  xlab = "Time", 
  ylab = "Survival Probability"
)

```

####Kaplan-Meier Analysis of Clusters by Weight Group
```{r}
#model by weight group within clusters
fit_weight_group_cluster <- survfit(Surv(overall_survival, deceased) ~ weight_group + Cluster, data = unq_patient_df)

#Kaplan-Meier plot
ggsurvplot(
  fit_weight_group_cluster, 
  data = unq_patient_df, 
  pval = TRUE, 
  risk.table = TRUE, 
  risk.table.col = "strata", 
  risk.table.height = 0.3,
  facet.by = "Cluster",
  title = "Kaplan-Meier Analysis of Clusters by Weight Group",
  xlab = "Time", 
  ylab = "Survival Probability"
)

```

####Kaplan-Meier Analysis of Clusters by Genetic Ancestry
```{r}
#model by genetic ancestry within clusters
fit_genetic_ancestry_cluster <- survfit(Surv(overall_survival, deceased) ~ GENETIC_ANCESTRY_LABEL + Cluster, data = unq_patient_df)

#Kaplan-Meier plot
ggsurvplot(
  fit_genetic_ancestry_cluster, 
  data = unq_patient_df, 
  pval = TRUE, 
  risk.table = TRUE, 
  risk.table.col = "strata", 
  risk.table.height = 0.3,
  facet.by = "Cluster",
  title = "Kaplan-Meier Analysis of Clusters by Weight Group",
  xlab = "Time", 
  ylab = "Survival Probability"
)

```




#Mutation ANalysis on Patients in Clusters to Find Mutational Patterns 
 

#DESeq2 Analysis on Expression Data to compare the clusters

```{r}
# prepare expression data for DESeq

de_RNA <- preprocessed_RNA
# Setup DESeq Object and run DESeq Pipeline 
colData <- clusters
countData <- de_RNA

# Set up Cluster numbers as factor for DESeq
colData$Cluster <-  as.factor(colData$Cluster)

# Create DESeq data set
dds <- DESeqDataSetFromMatrix(countData =  countData,
                              colData = colData,
                              design = ~ Cluster)

```

```{r}
dds <- DESeq(dds)
```

```{r}
dds

```
```{r}
#building DESeq results table
res <- results(dds, contrast = c("Cluster", "1", "2"))
res_df <- as.data.frame(res)

#summary 
summary(res)
```
```{r}
#Create Volcano Plot

# Function 1: Prepare DESeq2 results for plotting
# Convert DESeqResults object to data.frame
prepare_res_for_plot <- function(res) {
  res_df
  # rownames as a gene column
  res_df$hgnc_symbol <- rownames(res_df) 

  return(res_df)
}

# Function 2: Volcano plot function
volcplot <- function(data, padj_threshold = 0.05, log2Fold_threshold = 1, plot_title = 'Volcano Plot', plot_subtitle = NULL) {
  # Set the fold-change thresholds
  neg_log2fc <- -log2Fold_threshold
  pos_log2fc <- log2Fold_threshold

  # Replace NA values
  data$padj[is.na(data$padj)] <- 1
  data$log2FoldChange[is.na(data$log2FoldChange)] <- 0

  # Add log2fc_threshold column
  data$log2fc_threshold <- ifelse(
    data$log2FoldChange >= pos_log2fc & data$padj <= padj_threshold, 'up',
    ifelse(data$log2FoldChange <= neg_log2fc & data$padj <= padj_threshold, 'down', 'ns')
  )


  # Count up, down, and unchanged genes
  up_genes <- sum(data$log2fc_threshold == 'up')
  down_genes <- sum(data$log2fc_threshold == 'down')
  unchanged_genes <- sum(data$log2fc_threshold == 'ns')

  # Generate legend labels
  legend_labels <- c(
    paste0('Up: ', up_genes),
    paste0('Not significant: ', unchanged_genes),
    paste0('Down: ', down_genes)
  )

  # Calculate x-axis limits
  x_axis_limits <- ceiling(max(abs(data$log2FoldChange)))

  # Define plot colors
  plot_colors <- c(
    'up' = 'firebrick1',
    'ns' = 'gray',
    'down' = 'dodgerblue1'
  )

  # Create the plot
  plot <- ggplot(data) +
    geom_point(
      aes(x = log2FoldChange, y = -log10(padj), color = log2fc_threshold),
      alpha = 0.25,
      size = 1.5
    ) +
    geom_vline(xintercept = c(neg_log2fc, pos_log2fc), linetype = 'dashed') +
    geom_hline(yintercept = -log10(padj_threshold), linetype = 'dashed') +
    scale_x_continuous(
      'log2(FC)',
      limits = c(-x_axis_limits, x_axis_limits)
    ) +
    scale_color_manual(
      values = plot_colors,
      labels = legend_labels
    ) +
    labs(
      color = paste('log2Fold:', log2Fold_threshold, ', padj', expression("\u2264"), padj_threshold),
      title = plot_title,
      subtitle = plot_subtitle
    ) +
    theme_bw(base_size = 14) +
    theme(
      aspect.ratio = 1,
      axis.text = element_text(color = 'black'),
    )
  
  return(plot)
}

# Generate the volcano plot
volcplot(
  data = res_df,
  padj_threshold = 0.05,
  log2Fold_threshold = 1,
  plot_title = "Volcano Plot for DESeq2 Results",
  plot_subtitle = "Comparison: hoxa1_kd vs control_sirna"
)
```


```{r}
# retrieve all significant genes

genes.significant <- which(!is.na(res$padj) & 
                 res$padj < 0.05 & 
                 abs(res$log2FoldChange) > 1)

resSig <- subset(res, padj < 0.05)
# Get the indices for top 20 upregulated and downregulated genes
genes.top.upreg <- order(resSig$log2FoldChange,decreasing = TRUE)[1:20]
genes.top.downreg <- order(resSig$log2FoldChange,decreasing = FALSE)[1:20]

# Bind the two lists of genes into a single vector
genes.top <- c(genes.top.upreg, genes.top.downreg)

```


#Pathway Analysis
```{r}
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]


foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)


# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)


## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids


pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")

```



```{r}
#KEGG data
data(kegg.sets.hs)
data(sigmet.idx.hs)
kegg.sets.hs <- kegg.sets.hs[sigmet.idx.hs]  # For signaling and metabolic pathways

#function to run pathway analysis for a cluster
run_pathway_analysis_top5 <- function(cluster_name, log2_fold_changes) {
  #GAGE pathway analysis
  keggres <- gage(log2_fold_changes, gsets = kegg.sets.hs)
  
  #top pathways (upregulated and downregulated)
  top_up_pathways <- head(keggres$greater[order(keggres$greater[, "q.val"]), ], 5)
  top_down_pathways <- head(keggres$less[order(keggres$less[, "q.val"]), ], 5)

  
  #top 5 upregulated pathways
  if (!is.null(top_up_pathways) && nrow(top_up_pathways) > 0) {
    for (i in 1:min(5, nrow(top_up_pathways))) {
      top_pathway_id <- substr(rownames(top_up_pathways)[i], 1, 8)  #KEGG pathway ID
      message("Visualizing Upregulated Pathway ", i, ": ", rownames(top_up_pathways)[i])
      pathview(
        gene.data = log2_fold_changes, 
        pathway.id = top_pathway_id, 
        species = "hsa", 
        out.suffix = paste0("Cluster_", cluster_name, "_Upregulated_", i)
      )
    }
  }
  
  #top 5 downregulated pathways
  if (!is.null(top_down_pathways) && nrow(top_down_pathways) > 0) {
    for (i in 1:min(5, nrow(top_down_pathways))) {
      top_pathway_id <- substr(rownames(top_down_pathways)[i], 1, 8)  #KEGG pathway ID
      message("Visualizing Downregulated Pathway ", i, ": ", rownames(top_down_pathways)[i])
      pathview(
        gene.data = log2_fold_changes, 
        pathway.id = top_pathway_id, 
        species = "hsa", 
        out.suffix = paste0("Cluster_", cluster_name, "_Downregulated_", i)
      )
    }
  }
}

#subset the expression data for each cluster
cluster_1_data <- preprocessed_RNA[, colnames(preprocessed_RNA) %in% clust1_IDs]
cluster_2_data <- preprocessed_RNA[, colnames(preprocessed_RNA) %in% clust2_IDs]

#calculate log2 fold changes for each cluster
cluster_1_log2fc <- rowMeans(cluster_1_data) - rowMeans(preprocessed_RNA)
cluster_2_log2fc <- rowMeans(cluster_2_data) - rowMeans(preprocessed_RNA)

#perform pathway analysis
run_pathway_analysis_top5("Cluster_1", cluster_1_log2fc)
run_pathway_analysis_top5("Cluster_2", cluster_2_log2fc)


```

