---
title: "Scrap"
author: "Jung Kyu Yoon"
date: "2024-12-07"
output: html_document
---






Mutation data attempt:

_______________________________________________________________________________________________________________________________________
Mutation data:
1. Build a gene-patient matrix that contains mutation data. Within this matrix, "1" represents whether there is a mutation in a given gene and given patient and "0" represents no mutation. Note, when building this matrix, you can take into account the type of mutations (e.g., synonymous, non-synonymous, etc) 
2. Pick the top 20 or most frequently mutated genes from this matrix.  
3. Perform clustering on the resulting mutation matrix Perhaps try ward and ward.D2 linkage methods if you see narrow clusters.
4. Take the clusters of patients and do (a) DE analysis on all genes (not just top 500), (b) do pathway analysis on the DE genes that pass the adjusted p-value threshold (not just the top 20 genes) to see if specific pathways are hit, (c) see if outcome data between patient clusters is different, (d) see if any other clinical data is different when you compare patient clusters, (e) explore the correlation between the patient clusters that you have got from mutation and expression data - how much overlap do you see between the patient clusters?
_______________________________________________________________________________________________________________________________________

1. Build a gene-patient matrix that contains mutation data. Within this matrix, "1" represents whether there is a mutation in a given gene and given patient and "0" represents no mutation. Note, when building this matrix, you can take into account the type of mutations (e.g., synonymous, non-synonymous, etc)
```{r}
#filtered_mutations for processing without modifying the original fitlered data
filtered_mutations_copy <- filtered_mutations

#specific mutation types
valid_mutation_types <- c("missense_variant", "nonsense_variant", "frameshift_variant")
filtered_mutations_processed <- filtered_mutations_copy[filtered_mutations_copy$Consequence %in% valid_mutation_types, ]

#get unique Gene (Hugo_Symbol) and Patient (Tumor_Sample_Barcode) pairs
gene_patient_data <- unique(filtered_mutations_processed[, c("Hugo_Symbol", "Tumor_Sample_Barcode")])

#binary column ( 1= mutation )
gene_patient_data$mutated <- 1

#gene-patient matrix
gene_patient_matrix <- reshape2::dcast(
  gene_patient_data,
  Hugo_Symbol ~ Tumor_Sample_Barcode,
  value.var = "mutated",
  fill = 0
)
head(gene_patient_matrix)


```

2. Pick the top 20 or most frequently mutated genes from this matrix.  
```{r}
#total number of mutations for each gene
gene_patient_numeric <- as.matrix(gene_patient_matrix[, -1])  
rownames(gene_patient_numeric) <- gene_patient_matrix$Hugo_Symbol

#number of mutations across patients
gene_mutation_counts <- rowSums(gene_patient_numeric)

#mutation counts to the matrix
gene_patient_matrix$Mutation_Counts <- gene_mutation_counts

#sort by mutation counts in descending order
sorted_genes <- gene_patient_matrix[order(-gene_patient_matrix$Mutation_Counts), ]

#top 20 most frequently mutated genes
top_20_genes <- sorted_genes[1:20, ]

top_20_genes


```

3. Perform clustering on the resulting mutation matrix Perhaps try ward and ward.D2 linkage methods if you see narrow clusters.
```{r}

#mutation matrix for clustering (top 20 genes only)
top_20_matrix <- as.matrix(top_20_genes[, -c(1, ncol(top_20_genes))])
rownames(top_20_matrix) <- top_20_genes$Hugo_Symbol

#hierarchical clustering
dist_matrix <- dist(top_20_matrix, method = "binary")

#clustering with Ward
hc_ward <- hclust(dist_matrix, method = "ward.D")

#clustering with Ward.D2
hc_ward_d2 <- hclust(dist_matrix, method = "ward.D2")

#Plot dendrograms
par(mfrow = c(1, 2))  # Set up a side-by-side layout
plot(hc_ward, main = "Hierarchical Clustering (Ward)", xlab = "", sub = "", cex = 0.8)
plot(hc_ward_d2, main = "Hierarchical Clustering (Ward.D2)", xlab = "", sub = "", cex = 0.8)

```




4a. Take the clusters of patients and do (a) DE analysis on all genes (not just top 500)

```{r}
library(DESeq2)
library(reshape2)

# Reuse the hierarchical clustering result from Step 3
# Ensure cluster_assignments is created from the hierarchical clustering
cluster_assignments <- cutree(hc_ward_d2, k = 2)  # Replace hc_ward with hc_ward_d2 if preferred
```


```{r}
# Align patients with cluster assignments
valid_patients <- intersect(colnames(gene_patient_matrix)[-1], names(cluster_assignments))  # Exclude the first column (Hugo_Symbol)

cluster_assignments
```


```{r}
# Ensure valid patients exist
if (length(valid_patients) == 0) {
  stop("No valid patients found. Check alignment of patient IDs between gene_patient_matrix and cluster_assignments.")
}

# Subset the mutation matrix for valid patients
mutation_matrix <- as.matrix(gene_patient_matrix[, valid_patients, drop = FALSE])
rownames(mutation_matrix) <- gene_patient_matrix$Hugo_Symbol

# Ensure mutation_matrix contains numeric values
mutation_matrix <- apply(mutation_matrix, 2, as.numeric)

# Prepare colData with cluster assignments for valid patients
col_data <- data.frame(cluster = as.factor(cluster_assignments[valid_patients]))
rownames(col_data) <- valid_patients

# Validate alignment between mutation_matrix and colData
if (!all(colnames(mutation_matrix) == rownames(col_data))) {
  stop("Mismatch between mutation_matrix columns and colData rows. Check patient ID alignment.")
}

# Create DESeqDataSet
dds <- DESeqDataSetFromMatrix(
  countData = mutation_matrix,
  colData = col_data,
  design = ~ cluster
)

# Run DESeq2 pipeline
dds <- DESeq(dds)

# Extract results
res <- results(dds, alpha = 0.05)

# Subset significant results
resSig <- subset(res, padj < 0.05)

# View significant results
head(resSig)


```

```{r}
# Create DESeq2 dataset
dds <- DESeqDataSetFromMatrix(
  countData = mutation_matrix,
  colData = col_data,
  design = ~ cluster
)

# Run DESeq2 pipeline
dds <- DESeq(dds)

# Extract results
res <- results(dds, alpha = 0.05)
summary(res)

```




_________________________________________________________________________________________________________________________________
Expression data:
1. Normalize expression data using DEseq2. This was explained and included in the lab document. 
2. Pick the top n (e.g., 500, 1000, ...) most variable genes. This could be done before step 1 or after step 1. 
3. Perform clustering on the resulting expression matrix. Perhaps try ward and ward.D2 linkage methods if you see narrow clusters.
4. Take the clusters of patients and do (a) DE analysis on all genes (not just top 500), (b) do pathway analysis on the DE genes that pass the adjusted p-value threshold (not just the top 20 genes) to see if specific pathways are hit, (c) see if outcome data between the patient clusters is different, (d) see if any other clinical data is different when you compare patient clusters, (e) see if patient clusters are enriched with any type of mutation data.   
_________________________________________________________________________________________________________________________________
```{r}

```








